%!TEX root = ../template.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% acknowledgements.tex
%% NOVA thesis document file
%%
%% Text with acknowledgements
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\typeout{NT FILE acknowledgements.tex}%

\begin{ntacknowledgements}

I start by thanking Mário, my supervisor, for all of the time and work put into this thesis so that we could create something new, useful, and (in my biased perspective) wonderful. 
At the beginning, I doubted if I could understand all of those complex papers and make a contribution to science, as little as the impact would be.
However, confidence quickly built up when I put \emph{as mãos na massa} and I had a much better time than I ever imagined I could developing a master's thesis.
And, because of that, I blame Mário for having chosen a theme that fit me so well :)

I am also grateful for all the help that Ion gave me during this last year.
He and all of my lab colleagues were definitely a motivating force that drove me to want to do things as incredible as they did.
Furthermore, I would also like to thank professor Simão Melo de Sousa for being the rapporteur for this thesis.

Finally, I want to thank Faculdade de Ciências e Tecnologia and Universidade Nova de Lisboa for making all of this possible and for having an incredible culture, colleagues and professors that taught me so much in such a welcoming way.
And also, for being located in a place that is so amazingly well balanced for my taste that I proudly consider it my second favorite town in the world, only behind the place where I was born and raised.

\iffalse
Acknowledgments are personal text and should be a free expression of the author.
However, without any intention of conditioning the form or content of this text, I would like to add that it usually starts with academic thanks (instructors, etc.); then institutional thanks (Research Center, Department, Faculty, University, FCT / MEC scholarships, etc.) and, finally, the personal ones (friends, family, etc.).
But I insist that there are no fixed rules for this text, and it must, above all, express what the author feels.
\fi

\end{ntacknowledgements}
